Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-05-28 data.qmd

group_and_plot(
  df=contra_by_t,
  column="t_stage",
  axes=left,
  colors=[COLORS["blue"], COLORS["orange"]],
)
```

@@

```{python}
#| label: fig-data-upset-ipsi
#| fig-cap: Upset plot of the ipsilateral lymphatic involvement patterns [@lex_upset_2014]. Here, the horizontal bars depict the overall prevalence of LNL involvement. I.e., LNL II is the only level that is involved in more than 500 patients. The vertical bar plots depict joint involvements. E.g., the fourth bar from the left depicts how often the LNLs II, III, and IV are jointly involved while level I and V are not. This is indicated by the light and dark shaded dots in the matrix underneath.
from typing import Literal
import upsetplot

@@

axes = plot_involvement_upset(
  df=raw["max_llh", "ipsi"],
  lnls=[tpl[-1] for tpl in get_lnl_cols("ipsi")],
)
plt.show()
```

@@

```{python}
#| label: fig-data-upset-contra
#| fig-cap: Upset plot of the contralateral lymphatic involvement patterns [@lex_upset_2014]. Here, the horizontal bars depict the overall prevalence of LNL involvement. I.e., LNL II is the only level that is involved in more than 100 patients. The vertical bar plots depict joint involvements. E.g., the thirs bar from the left depicts how often the LNLs II, and III are jointly involved while level I, IV, and V are not. This is indicated by the light and dark shaded dots in the matrix underneath.

axes = plot_involvement_upset(
  df=raw["max_llh", "contra"],
  lnls=[tpl[-1] for tpl in get_lnl_cols("contra")],
)
plt.show()
```

>>>>

group_and_plot(
  df=contra_by_t,
  column="t_stage",
  axes=left,
  colors=[COLORS["blue"], COLORS["orange"]],
)

plt.savefig(paths.figure_dir / "fig-data-strat.png", dpi=400)
plt.show()
```

@@

```{python}
#| label: fig-data-upset-ipsi
#| fig-cap: Upset plot of the ipsilateral lymphatic involvement patterns [@lex_upset_2014]. Here, the horizontal bars depict the overall prevalence of LNL involvement. I.e., LNL II is the only level that is involved in more than 500 patients. The vertical bar plots depict joint involvements. E.g., the fourth bar from the left depicts how often the LNLs II, III, and IV are jointly involved while level I and V are not. This is indicated by the light and dark shaded dots in the matrix underneath.
from typing import Literal
import upsetplot
from scripts import paths

@@

axes = plot_involvement_upset(
  df=raw["max_llh", "ipsi"],
  lnls=[tpl[-1] for tpl in get_lnl_cols("ipsi")],
)

plt.savefig(paths.figure_dir / "fig-data-upset-ipsi.png", dpi=400)
plt.show()
```

@@

```{python}
#| label: fig-data-upset-contra
#| fig-cap: Upset plot of the contralateral lymphatic involvement patterns [@lex_upset_2014]. Here, the horizontal bars depict the overall prevalence of LNL involvement. I.e., LNL II is the only level that is involved in more than 100 patients. The vertical bar plots depict joint involvements. E.g., the thirs bar from the left depicts how often the LNLs II, and III are jointly involved while level I, IV, and V are not. This is indicated by the light and dark shaded dots in the matrix underneath.
from scripts import paths

axes = plot_involvement_upset(
  df=raw["max_llh", "contra"],
  lnls=[tpl[-1] for tpl in get_lnl_cols("contra")],
)

plt.savefig(paths.figure_dir / "fig-data-upset-contra.png", dpi=400)
plt.show()
```

### 2024-05-28 manuscript.qmd

![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region. Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables. Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step. The orange arcs represent the sensitivity and specificity of the observational modality (e.g. CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=60%}

Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019]. We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2020] to allow an intuitive inclusion of T-category into the predictions. We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.

>>>>

![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region. Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables. Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step. The orange arcs represent the sensitivity and specificity of the observational modality (e.g. CT, MRI, pathology, ...).](static/small-graph.png){#fig-small-graph width=50%}

Our first model to predict the lymphatic progression of HNSCC was introduced using Bayesian networks [@pouymayou_bayesian_2019]. We subsequently extended this work to a hidden Markov model (HMM) [@ludwig_hidden_2021] to allow an intuitive inclusion of T-category into the predictions. We will briefly summarize this HMM's formalism before building on it to include the contralateral spread in @sec-ext-to-contra.
